### 2015-05-06 doc/thesis_puc/typedlua.tex
Fixing typos

Another goal of Typed Lua is to be backwards compatible with Lua.
This means that any Lua code is valid Typed Lua code.
To be backwards compatible with Lua, the syntactic extensions introduced
by Typed Lua does not include new reserved words.
Appendix \ref{app:syntax} presents the complete syntax of Typed Lua
in extended BNF.

@@

In this example, the compiler uses local type inference to assign the
type \texttt{integer} to the local variable \texttt{x} and to the
return type of the local function \texttt{succ}, making this example
compile without any warnings.
Local type inference always use the most general type.
In this example, the compiler does not use the literal type \texttt{7},
instead of the base type \texttt{integer}, because this would generate
a warning when we try to assign other integer value to the variable \texttt{x}.
Still, programmers can use literal types in type annotations if
they need a variable that has a very specific type.
In Section \ref{sec:tables} we will see in more detail that literal types
are essential to type Lua tables.

@@

    local function distance (x, y)
      return abs(x - y)
    end
\end{verbatim}

@@

Lua has first-class functions, but they have some peculiarities.
First, the number of arguments of a function call does
not need to match the arity of the function declaration, as Lua silently
drops extra arguments after evaluating them, or uses \texttt{nil} to
replace missing arguments.
Second, functions can return any number of values,
and this number of returned values may not be statically known.
Third, Lua has multiple assignment, and the semantics
of argument passing is the same of the multiple assignment, that is,
calling a function is like doing a multiple assignment where the
left-hand side is the parameter list and the right-hand side is the
argument list.

@@

Even though these operation modes affect function calls, they do not
affect other multiple assignment.
For instance,
\begin{verbatim}
    local x:integer, y:integer = 1, 2, 3
\end{verbatim}
compiles without any warnings in both operation modes.
This means that the left-hand side of this local declaration has the
tuple type \texttt{(integer, integer, value*)}, while the expression
list in the right-hand side of this local declaration has the
tuple type \texttt{(1, 2, 3, nil*)} in both operation modes.

@@

Typed Lua always includes \texttt{nil*} in the end of the type
of an expression list that does not end in a varidic type.
This behavior preserves the semantics of Lua on replacing missing values,
and it is necessary when we omit optional parameters in a function call.
We will discuss optional parameters in the next section.

@@

The table type
$\texttt{\{l}{_\texttt{1}}\texttt{:t}_{\texttt{1}}
\texttt{, ..., l}{_\texttt{n}}\texttt{:t}_{\texttt{n}}\texttt{\}}$
represents the type of a record that maps literal types
$\texttt{l}_{\texttt{i}}$\texttt{, ..., }$\texttt{l}_{\texttt{n}}$
to values of types $\texttt{t}_{\texttt{i}}$\texttt{, ..., }$\texttt{t}_{\texttt{n}}$.
Most programming languages treat records as maps from names
to types, but we chose to use literal types instead of
names due to the statistics results that we obtained about
the usage of the table constructor, which we discussed in
Section \ref{sec:statistics}.
This means that we can use the syntax of records to type a
list that has fixed elements, like the variable \texttt{days} from
the previous example.

@@

    print(greeter(user1))             --> Goodbye Lewis Reed 
    print(greeter(user2))             --> compile-time warning
    print(greeter(user3))             --> Goodbye Lou Reed
    print(greeter(user4))             --> compile-time warning
\end{verbatim}

@@

Even though the interface declaration may look redundant due to
the type alias declaration, it has a more convenient syntax
for declaring table types that express records.
For this reason, it worth having two different constructs,
one specifically for records and another for more general types.

@@

Now we can write a program that declares a list, inserts some
elements in this list, and then traverses the list printing
each list element:
\begin{verbatim}
    local l:Element?

@@

The Typed Lua compiler searches for a
description file when it cannot not find the respective Typed Lua file
that is the argument of \texttt{require}.
In this example, the call to \texttt{require} assigns to the local
\texttt{m} the table type that the description file of the \texttt{lmd5}
library exports. 
Thus, the compiler raises a compile-time warning in the fourth line,
as the function \texttt{clone} expects a value of type
\texttt{md5\string_context} instead of a value of type \texttt{string}.

>>>>

Another goal of Typed Lua is to be backwards compatible with Lua.
This means that any Lua code is valid Typed Lua code.
To be backwards compatible with Lua, the syntactic extensions introduced
by Typed Lua do not include new reserved words.
Appendix \ref{app:syntax} presents the complete syntax of Typed Lua
in extended BNF.

@@

In this example, the compiler uses local type inference to assign the
type \texttt{integer} to the local variable \texttt{x} and to the
return type of the local function \texttt{succ}, making this example
compile without any warnings.
Local type inference always uses the most general type.
In this example, the compiler does not use the literal type \texttt{7},
instead of the base type \texttt{integer}, because this would generate
a warning when we try to assign other integer value to the variable \texttt{x}.
Still, programmers can use literal types in type annotations if
they need a variable that has a very specific type.
In Section \ref{sec:tables} we will see in more detail that literal types
are essential to type Lua tables.

@@

    local function distance (x, y)
      return absolute(x - y)
    end
\end{verbatim}

@@

Lua has first-class functions, and they have some peculiarities.
First, the number of arguments of a function call does
not need to match the arity of the function declaration, as Lua silently
drops extra arguments after evaluating them, or uses \texttt{nil} to
replace missing arguments.
Second, functions can return any number of values,
and this number of returned values may not be statically known.
Third, Lua has multiple assignment, and the semantics
of argument passing is the same of the multiple assignment, that is,
calling a function is like doing a multiple assignment where the
left-hand side is the parameter list and the right-hand side is the
argument list.

@@

Even though these operation modes affect function calls, they do not
affect other multiple assignments.
For instance,
\begin{verbatim}
    local x:integer, y:integer = 1, 2, 3
\end{verbatim}
compiles without any warnings in both operation modes.
This means that the left-hand side of this local declaration has the
tuple type \texttt{(integer, integer, value*)}, while the expression
list in the right-hand side of this local declaration has the
tuple type \texttt{(1, 2, 3, nil*)} in both operation modes.

@@

Typed Lua always includes \texttt{nil*} in the end of the type
of an expression list that does not end in a variadic type.
This behavior preserves the semantics of Lua on replacing missing values,
and it is necessary when we omit optional parameters in a function call.
We will discuss optional parameters in the next section.

@@

The table type
$\texttt{\{l}{_\texttt{1}}\texttt{:t}_{\texttt{1}}
\texttt{, ..., l}{_\texttt{n}}\texttt{:t}_{\texttt{n}}\texttt{\}}$
represents the type of a record that maps literal types
$\texttt{l}_{\texttt{i}}$\texttt{, ..., }$\texttt{l}_{\texttt{n}}$
to values of types $\texttt{t}_{\texttt{i}}$\texttt{, ..., }$\texttt{t}_{\texttt{n}}$.
Most programming languages treat records as maps from names
to types, but we chose to use literal types instead of
names due to the statistics results that we obtained about
the usage of the table constructor, which we discussed in
Section \ref{sec:statistics}.
This means that we can use the syntax of records to type a
list that has a fixed number of elements,
like the variable \texttt{days} from the previous example.

@@

    print(byebye(user1))             --> Goodbye Lewis Reed 
    print(byebye(user2))             --> compile-time warning
    print(byebye(user3))             --> Goodbye Lou Reed
    print(byebye(user4))             --> compile-time warning
\end{verbatim}

@@

Even though the interface declaration may look redundant due to
the type alias declaration, it has a more convenient syntax
for declaring table types that express records.
For this reason, it is worth having two different constructs,
one specifically for records and another for more general types.

@@

Now we can write a program that declares a list, inserts some
elements in it, and then traverses the list printing
each element:
\begin{verbatim}
    local l:Element?

@@

The Typed Lua compiler searches for a
description file when it cannot find the respective Typed Lua file
that is the argument of \texttt{require}.
In this example, the call to \texttt{require} assigns to the local
\texttt{m} the table type that the description file of the \texttt{lmd5}
library exports. 
Thus, the compiler raises a compile-time warning in the fourth line,
as the function \texttt{clone} expects a value of type
\texttt{md5\string_context} instead of a value of type \texttt{string}.
